\thispagestyle{empty}
\begin{spacing}{1.0}
\chapter*{Acknowledgments}

Firstly, I would like to dedicate this work to my mother, Andrea. Mom, no matter how hard I try, I know that I will never be able to express my gratitude in words the way I would like. Without you, I wouldn't even be here today. I would not even have finished this cycle, so important in my life. Thank you for being my support since I entered college. Thank you for always being willing to take care of me, especially when I'm not well, even though I'm an adult. Thank you for always being available for a few minutes (which often turn into hours) of conversation. Thank you for always giving me the best advice in the world. Thank you for believing in me when I often can't believe it myself. Thank you for always being with me, in the highs and, especially, in the lows. Thank you for sharing your life with me and for allowing me to share mine with you. Thank you for being peace and calm in the midst of so much storm. Without you, I would be nothing. Thank you for accepting me the way I am and at the same time always encouraging and motivating me to be better than I was yesterday. Thank you for loving me so intensely and in such a sublime way that I have no doubt that it transcends incarnations. I love you deeply with all the strength of my being. This thesis is for you, mom.

To my father, Alvaro, thank you for so much. Firstly, thank you for giving me the chance to study at excellent institutions and for supporting me and giving me the opportunity to focus on my studies during the period I was at university. Thank you for being there to visit me, even though you are so far away from me. Thank you for loving me and trying to understand me the way I am. Thank you for never letting me lack anything in this life. You are, without a shadow of a doubt, an exceptional father. I love you so much.

I would like to thank Professor Leopoldo Teixeira not only for introducing me to the area of Compilers and Programming Languages, but mainly for accepting to be my advisor for the preparation of my undergraduate thesis. Professor, thanks to your classes, your excellent teaching skills and your good humor, I was able to find an area within computer science with which I could truly identify. I would like to thank you with all my heart for your help, advice and patience in talking to me during these last few months. I believe that, fortunately, we were able to build a true friendship and not be restricted to a mere formal relationship between student and professor. I look forward to working together again in a near future. I wish you and your family all the best.

I would also like to dedicate this work, in which I put my soul and my heart, to my grandmother, Maria Dolores, who, unfortunately, is no longer with us. We had a relationship that was so fleeting, but at the same time so intense and full of meaning that it seemed like decades. I don't believe in coincidences, so I see the fact that my defense was scheduled on your birthday as a superior opportunity to honor you with this more than special gift. I would like to say that I miss you every day and that it really saddens me that you are not here, physically present, to see your grandson's defense. However, what comforts me is the feeling that wherever you are, I know you are very proud of the person I have become.

Obviously, I couldn't forget my sweet and dear Laura. How could I forget someone who has been following me for a while with so much admiration, love, affection and zeal? Laura, I thank you for being so understanding with me during this phase of my life that has been so full of surprises. Thank you for always trying to encourage and support me. Thank you for being present in my life, even when things aren't going so well. Thank you for loving me and accepting me the way I am, without taking anything away. Never forget that, thanks to you, this journey could be easier. I love you.

I would also like to thank my colleagues who enrolled into the Computer Engineering program with me back in 2018, especially: Daniel Perazzo, Fernando Macedo, Lucas Ambrosio, Matheus Teotônio, Riei Joaquim, Victor Hugo, Victor Ximenes and Zênio Ângelo. Thanks to each one of you, this journey was less heavy and also more funny than it would probably be.

Talking about colleagues, I think it would not be fair of my part if I didn't acknowledge the larger impact that José Carlos da Silva Cruz, Pedro Nogueira Coutinho and Zilde Souto Maior Neto had in my life. After all, different people have different impacts in our lives. I am very grateful that we were able to build strong bonds through these last 6 years. It has been a wild ride.

José, thank you for helping me in the professional aspect of my life. You always had my back, not only when I was in need of guidance about career opportunities, but also when I just needed a friend to look over me in problematic situations. You also taught me a lot about altruism, hard work and perseverance. To me, you are a guardian angel that god put in my life. I wish you nothing but the very best in your life.

Pedro, we started our friendship very slowly but I think our own common problems in certain aspects of our lives made this process both faster and stronger. Thank you for always having the patience to listen to me and also offer advice and help when I needed. I feel very blessed that you trusted me to share your history, your life and your problems with me so I could help you whenever possible.

Zilde, you have nothing but my gratitude. I'd like to thank you for helping that poor guy that was struggling so much with the Introduction to Programming course back in 2018. Alias, I'd like to thank you for helping me whenever I needed during these 6 years that we spent together at UFPE. I'd also like to say that you have been motivating me to be a better person, since the day we first met. Your willingness to help others and your dedication to be a better professional had an immense positive influence in me.

To those who came before me, in special: José Victor Silva Cruz, Lucas Santana and Marcos Oliveira, I have always admired the three of you during the few years that we were interacting with each other at UFPE. Thanks to this brief acquaintanceship, I was able to become not only a better student, but also a better developer and person. 

I would also like to thank Cintia, my psychologist, for doing an incredible work with me since 2022. Her help and advice guided me in order to become a better version of myself during these last 2 years.

Lastly, even though he probably won't ever see this, I'd like to thank Mr. Noriaki Kubo. Thank you for creating the work of fiction that marked my life so much, especially in its current stage, to the point that I named the work presented here after it.
    
\clearpage
\end{spacing}